Metric Spaces with Expensive Distances

### 2019-11-26 paper/draft_socg.tex
add citations

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Our experiments with Earth Mover distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 3, saving hundreds of hours of machine time,
even though the images and diagrams in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

For experiments with images, we used the German traffic sign
dataset and the code for computing the Earth Mover distance
by Gary Dornan.
We randomly selected 250 images from the benchmark.
The first two columns of \cref{tbl:emd_gtsrb}
compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
For $10 \%$ approximation, we can reduce
the running time by more than a factor of $2$,
saving $98$ hours of computation time.
For $20 \%$ approximation,
we speed up the computations by a factor of 3.

@@

\myparagraph{Earth Mover distance}
Let $\mu_P$ and $\mu_Q$
be two probability measures supported
on finite sets $P = \{p_1, \dots, p_n \}$
and $Q = \{ q_1, \dots, q_m \}$.
An $m \times n$ matrix $C$
is called a \textit{transportation plan},
if it satisfies the following three conditions:
\begin{enumerate}
    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
\end{enumerate}
%By definition, the \textit{cost} of transportation plan $C$
Let $\dist(\cdot, \cdot)$ denote the Euclidean
distance in plane.
The earth mover distance between $\mu_P$ and $\mu_Q$
is defined as
\[
    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j)$,
\]
where $\inf$ is taken over all transportation
plans $C$. Intuitively, $c_{i,j}$ is the
mass concentrated at the point $p_i$ that
is sent to the point $q_j$, and the earth mover
distance is the smallest total amount of work
needed to transform the measure $\mu_P$
into the measure $\mu_Q$.

>>>>

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of images
equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Our experiments with the Earth Mover's distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 2 to 5, saving hundreds of hours of machine time,
even though the inputs in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

For experiments with images, we used the German traffic sign
dataset \cite{gtsrb} and the code for computing the Earth Mover distance
by Doran \cite{doran_pyemd}.
We randomly selected 250 images from the benchmark.
The first two columns of \cref{tbl:emd_gtsrb}
compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
For $10 \%$ approximation, we can reduce
the running time by more than a factor of $2$,
saving $98$ hours of computation time.
For $20 \%$ approximation,
we speed up the computations by a factor of 3.

@@

\myparagraph{Earth Mover distance}
Let $\mu_P$ and $\mu_Q$
be two probability measures supported
on finite sets $P = \{p_1, \dots, p_n \}$
and $Q = \{ q_1, \dots, q_m \}$.
An $m \times n$ matrix $C$
is called a \textit{transportation plan},
if it satisfies the following three conditions:
\begin{enumerate}
    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
\end{enumerate}
%By definition, the \textit{cost} of transportation plan $C$
Let $\dist(\cdot, \cdot)$ denote the Euclidean
distance in plane.
The earth mover distance between $\mu_P$ and $\mu_Q$
is defined as
\[
    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j),
\]
where $\inf$ is taken over all transportation
plans $C$. Intuitively, $c_{i,j}$ is the
mass concentrated at the point $p_i$ that
is sent to the point $q_j$, and the earth mover
distance is the smallest total amount of work
needed to transform the measure $\mu_P$
into the measure $\mu_Q$.
